---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data


```{r, echo=TRUE}
## Loading packages
library(dplyr)
library(ggplot2)
library(data.table)

## Unpacking data
unzip('activity.zip', c("activity.csv"))
data <- read.csv("activity.csv")
data<-data.table(data)

## Preparing data for analysis
data_grouped<-data[,mean(steps,na.rm=T),by=date]
```


## What is mean total number of steps taken per day?

### Plotting Histrogram

```{r, echo=TRUE}
hist(data_grouped$V1,col="red",xlab="",main="Histogram of Steps")

```

### Calculation of mean and median
```{r, echo=TRUE}
mean<-mean(data_grouped$V1,na.rm=T)
median<-median(data_grouped$V1,na.rm=T)
```

The mean is `r mean` and the median is `r median`.

## What is the average daily activity pattern?

```{r, echo=TRUE}
avg_data<-data[,mean(steps,na.rm=T),by=interval]
plot(avg_data$interval,avg_data$V1,xlab="Interval",type="l", ylab="Average Steps",
     main="Interval Plotted Against Average Number of Steps")
```

```{r, echo=TRUE}
max_steps <- filter(avg_data, avg_data$V1==max(avg_data$V1))$interval
```

The interval containing the max number of averaged steps is `r max_steps`

## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
