---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

## Loading and preprocessing the data

```{r, echo=TRUE}
## Loading packages
library(dplyr)
library(ggplot2)
library(data.table)

## Unpacking data
unzip('activity.zip', c("activity.csv"))
data <- read.csv("activity.csv")
data<-data.table(data)
```


## What is mean total number of steps taken per day?

### Plotting Histogram

```{r, echo=TRUE}
data_grouped<-data[,mean(steps,na.rm=T),by=date]
hist(data_grouped$V1,col="red",xlab="",main="Histogram of Steps")

```

### Calculation of mean and median

```{r, echo=TRUE}
mean<-mean(data_grouped$V1,na.rm=T)
median<-median(data_grouped$V1,na.rm=T)
```

The mean is `r mean` and the median is `r median`.

## What is the average daily activity pattern?

```{r, echo=TRUE}
avg_data<-data[,mean(steps,na.rm=T),by=interval]
plot(avg_data$interval,avg_data$V1,xlab="Interval",type="l", ylab="Average Steps",
     main="Interval Plotted Against Average Number of Steps")
```

```{r, echo=TRUE}
max_steps <- filter(avg_data, avg_data$V1==max(avg_data$V1))$interval
```

The interval containing the max number of averaged steps is `r max_steps`

## Imputing missing values

```{r, echo=TRUE}
na_count = sum(is.na(data))
```

There are `r na_count` missing values in the dataset.

Missing values of steps will be replaced by the mean of the corresponding interval. A new dataset will be created that
is equal to the original dataset but with missing values filled in.

```{r, echo=TRUE}
joined_data <- left_join(data, avg_data, by="interval")
joined_data$steps[is.na(joined_data$steps)] <- joined_data$V1[is.na(joined_data$steps)]
joined_data$V1 <- NULL
filled_data <- data.table(joined_data)
```

Next, we'll create a histogram with the filled in dataset according to the method of the histogram made earlier:

### Plotting Histogram

```{r, echo=TRUE}
data_grouped<-filled_data[,mean(steps,na.rm=T),by=date]
hist(data_grouped$V1,col="red",xlab="",main="Histogram of Steps")
```

### Calculation of mean and median

```{r, echo=TRUE}
mean_filled<-mean(data_grouped$V1,na.rm=T)
median_filled<-median(data_grouped$V1,na.rm=T)
```

The mean is `r mean_filled` and the median is `r median_filled`.

The historgram gives the same pattern for both datasets, althought the exact numbers diverge slightly.
The means are the same (which makes sense as we used the means to fill the missing values). The medians also diverge
slightly. Conclusively: the changes to the dataset by filling in the NA's can be considered negligble. However,
as more observations are almost always favoured when applying machine learning techniques, fillign in the blanks
is worth the trouble.

## Are there differences in activity patterns between weekdays and weekends?


